Biden to win the popular vote in the 2020 American presidential elections with 51% of the votes

### 2020-11-03 outputs/paper/paper.Rmd
Final push

---
title: "Biden to win the popular vote in the 2020 American presidential elections in a closely fought contest"
author: "Anne Collins, Jennifer Do, Andrea Therese Javellana, Wijdan Tariq"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: pdf_document
abstract: \noindent "The 2020 American presidential elections are the most discussed elections in the world due to their potential and profound impact on global affairs. In this paper, we aim to predict whether Biden or Trump will win the popular vote and by how much of a margin. We ran a logistic regression model using 5,127 observations from election survey data provided by Democracy Fund + UCLA Netscape. To provide more robust predictions, we utilize the multilevel regression with poststratification method using over 2 million observations from the American Community Survey (ACS) data. While the survey data suggests that Biden will get 52% of the popular vote compared to Trump's 48%, our post-stratification results suggest that the election will be even closer, with 51% predicted for Biden and 49% for Trump. We conclude the paper with a discussion of the benefits of using post-stratification on survey data to correct for potential sample biases and to get more reliable regression estimates."
indent: true
---

@@

In this paper, we aim to predict the outcome of the overall popular vote of the 2020 American presidential election using public opinion polling data. People have different views over whether political polling are reliable predictors of election outcomes, especially after the shock victory of President Trump in 2016 which most pollsters failed to predict. Gelman and Azari (2017) point to nineteen lessons learned from that election, most relevant to our paper of which is the lesson that one needs to be cautious of survey nonresponse bias. Kennedy et al. (2018) showed that a late swing in vote preference toward Trump, a failure to adjust for overrepresentation of college graduates who mostly favored Clinton, and a clear change in voter turnout from 2012 to 2016 were some of the main reasons why pre-election polls performed poorly in 2016. Moreover, the rise of populism in America and the resulting countermovement makes elections more unpredictable than ever (Inglehart and Norris, 2016), even with the explosion of data availability. Notwithstanding these issues, the use of public opinion polling data continues to be common practice and we make use of it in this paper. (BRIEFLY discuss datasets here).

We find that Biden/Trump is likely to win! (briefly discuss results here).

@@

Introduce R, R packages used, logistic regression and MRP methods briefly here as well as a quick note on results..briefly introduce which data we use and what key findings were...WHO IS GOING TO WIN??...give a sketch of the rest of the report)

@@

The figures below compare the variables of the ACS data after cleaning with the survey data. From the plots generated from the survey data, it was observed that the majority of respondents were aged between 30 to 44 years, and were split evenly between male and female respondents. The majority of respondents were observed to be employed at 57% of the respondent population, followed by respondents not actively in the work force at 34% (Figure 3). The majority of respondents were White (69%), followed by Native American (14%), and Black Americans (10%), and other races ranged between 1-3% of the respondent population (Figure 4). Based on the survey, a greater distribution of respondents were seen to be more likely to vote for Joe Biden (Democratic candidate) in the upcoming 2020 election compared to Donald Trump (Republican candidate), at 51% and 48% of respondents answering respectively (Figure 5). This may be due to respondent bias, as the Nationscape dataset itself was in partnership with the Democracy Fund, it is expected that a greater number of respondents who support the democratic party would respond.

@@

Before going into the details of the post-stratification we conducted, we will discuss what post-stratification is and when it can be used. Post-stratification is implemented after simpler random samples have been conducted. It is used to properly balance the representation of variables across the target population to gain more precise estimates and create greater confidence in inferences being made. Post-stratification involves classifying each member of a population into a single subgroup or strata to then calculate the probability sample from each stratum. Because of this, post-stratification cannot be used in studies with observations that overlap or are not clearly classified, as this may inaccurately reflect the population. Another challenge is to find definitive lists of variables for an entire population that fall in line with the variables collected for the original sample.

@@

@@

Insert comment here from Kennedy and Gelman (2019, p. 19) about cost/benefits of MRP.

“Given the bias in the survey, discussed in Data section, we used MRP to...... While this would address some aspects of the bias [discuss bits that it can address], it cannot address all of it. This includes [discuss bits that it cannot address].”

Our model assumed people would either vote for Trump or Biden. In reality, independent voters would also...      so...Multinomial logistic regression model would be more appropriate.

@@

*Papers about US election predictions and predictions in general*

@@

@@

@@

Kennedy, C., Blumenthal, M., Clement, S., Clinton, J. D., Durand, C., Franklin, C., McGeeney, K., Miringoff, L., Olson, K., Rivers, D., Saad, L. Witt, G. E., & Wlezien, C. (2018). An evaluation of the 2016 election polls in the United States. *Public Opinion Quarterly*, 82(1), 1-33.

@@

@@

*Methodology papers*

Alexander, R. (2020). Getting started with MRP. Teaching note.

Chang, L., & Krosnick, J. A. (2009). National surveys via RDD telephone interviewing versus the Internet: Comparing sample representativeness and response quality. *Public Opinion Quarterly*, 73(4), 641-678.

Gelman, A. (2007). Struggles with survey weighting and regression modeling. *Statistical Science*, 22(2), 153-164.

Gelman, A. (2008). Scaling regression inputs by dividing by two standard deviations. *Statistics in Medicine*, 27(15), 2865-2873.

Gelman, A., & Hill, J. (2006). *Data analysis using regression and multilevel/hierarchical models*. Cambridge University Press.

Ghitza, Y., & Gelman, A. (2013). Deep interactions with MRP: Election turnout and voting patterns among small electoral subgroups. *American Journal of Political Science*, 57(3), 762-776.

Ghitza, Y., & Gelman, A. (2020). Voter Registration Databases and MRP: Toward the Use of Large-Scale Databases in Public Opinion Research. *Political Analysis*, 28(4), 507-531.

Hainmueller, J., Mummolo, J., & Xu, Y. (2019). How much should we trust estimates from multiplicative interaction models? Simple tools to improve empirical practice. *Political Analysis*, 27(2), 163-192.

Hanretty, C. (2019). An introduction to multilevel regression and post-stratification for estimating constituency opinion. *Political Studies Review*, 1478929919864773.

Jackman, S., Ratcliff, S., & Mansillo, L. (2019). Small area estimates of public opinion: model-assisted post-stratification of data from voter advice applications. Working paper.

Kennedy, L. & Gabry, J. (2020). MRP with rstanarm. Available at: https://cran.r-project.org/web/packages/rstanarm/vignettes/mrp.html. 20 July 2020.

Kennedy, L., & Gelman, A. (2019). Know your population and know your model: Using model-based regression and poststratification to generalize findings beyond the observed sample. *arXiv preprint arXiv:1906.11323*.

Lauderdale, B. E., Bailey, D., Blumenau, J., & Rivers, D. (2020). Model-based pre-election polling for national and sub-national outcomes in the US and UK. *International Journal of Forecasting*, 36(2), 399-413.

Tausanovitch, C., & Warshaw, C. (2013). Measuring constituent policy preferences in congress, state legislatures, and cities. *The Journal of Politics*, 75(2), 330-342.

Wang, W., Rothschild, D., Goel, S., & Gelman, A. (2015). Forecasting elections with non-representative polls. *International Journal of Forecasting*, 31(3), 980-991.

Yeager, D. S., Krosnick, J. A., Chang, L., Javitz, H. S., Levendusky, M. S., Simpser, A., & Wang, R. (2011). Comparing the accuracy of RDD telephone surveys and internet surveys conducted with probability and non-probability samples. *Public Opinion Quarterly*, 75(4), 709-747.

*Packages used*

Auguie, Baptiste (2017). gridExtra: Miscellaneous Functions for “Grid” Graphics. R package version 2.3. https://CRAN.R-project.org/package=gridExtra

Di Lorenzo, Paolo (2020). usmap: US Maps Including Alaska and Hawaii. R package version 0.5.1. https://CRAN.R-project.org/package=usmap

Douglas Bates, Martin Maechler, Ben Bolker, Steve Walker (2015). Fitting Linear Mixed-Effects Models Using lme4. Journal of Statistical Software, 67(1), 1-48. doi:10.18637/jss.v067.i01.

Firke, Sam (2020). janitor: Simple Tools for Examining and Cleaning Dirty Data. R package  version 2.0.1. https://CRAN.R-project.org/package=janitor

@@

Wickham, Hadley and Dana Seidel (2020). scales: Scale Functions for Visualization. R package version 1.1.1. https://CRAN.R-project.org/package=scales

@@

>>>>

---
title: "Biden to win the popular vote in the 2020 American presidential elections with 51% of the votes"
author: "Anne Collins, Jennifer Do, Andrea Therese Javellana, Wijdan Tariq"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: pdf_document
abstract: \noindent "The 2020 American presidential elections are the most discussed elections in the world due to their potential and profound impact on global affairs. In this paper, we aim to predict whether Biden or Trump will win the popular vote and by how much of a margin. We ran a logistic regression model using 5,127 observations from election survey data provided by Democracy Fund + UCLA Netscape. To provide more robust predictions, we utilize the multilevel regression with poststratification method using over 2 million observations from the American Community Survey (ACS) data. While the survey data suggests that Biden will get 52% of the popular vote compared to Trump's 48%, our post-stratification results suggest that the election will be even closer, with 51% predicted for Biden and 49% for Trump. We conclude the paper with a discussion of the benefits of using post-stratification on survey data to correct for potential sample biases and to get more reliable regression estimates."
indent: true
---

@@

In this paper, we aim to predict the outcome of the overall popular vote of the 2020 American presidential election using public opinion polling data. People have different views over whether political polling are reliable predictors of election outcomes, especially after the shock victory of President Trump in 2016 which most pollsters failed to predict. Gelman and Azari (2017) point to nineteen lessons learned from that election, most relevant to our paper of which is the lesson that one needs to be cautious of survey nonresponse bias. Kennedy et al. (2018) showed that a late swing in vote preference toward Trump, a failure to adjust for overrepresentation of college graduates who mostly favored Clinton, and a clear change in voter turnout from 2012 to 2016 were some of the main reasons why pre-election polls performed poorly in 2016. Moreover, the rise of populism in America and the resulting countermovement makes elections more unpredictable than ever (Inglehart and Norris, 2016), even with the explosion of data availability. Notwithstanding these issues, the use of public opinion polling data continues to be common practice and we make use of it in this paper.

Our analysis suggests that Biden is expected to win the popular vote. The survey data suggests that Biden's vote will win 52% of the vote, whilst the post-stratification estimates his share to be 51% compared to Trump's 49%.

We conducted our data wrangling and analysis using the statistical language R (R Core Team, 2020) using a variety of packages including tidyverse (Wickham, 2019), ggthemes (Arnold, 2019), lme4 (Bates et al., 2015), stargazer (Hlavac, 2018), usmap (Di Lorenzo, 2020), haven (Wickham and Miller, 2019), and labelled (Larmarange, 2020). Other packages used to generate this PDF file include tint (Eddelbuettel and Gilligan, 2020), kableExtra (Zhu, 2020), tinytex (Yihui, 2019), knitr (Yihui, 2020), latexpdf (Bergsma, 2018), and rmarkdown (Allaire et al., 2020).

@@

@@

The figures below compare the variables of the ACS data after cleaning with the survey data. From the plots generated from the survey data, it was observed that the majority of respondents were aged between 30 to 44 years, and were split evenly between male and female respondents. The majority of respondents were observed to be employed at 57% of the respondent population, followed by respondents not actively in the work force at 34% . The majority of respondents were White (69%), followed by Native American (14%), and Black Americans (10%), and other races ranged between 1-3% of the respondent population. Based on the survey, a greater distribution of respondents were seen to be more likely to vote for Joe Biden (Democratic candidate) in the upcoming 2020 election compared to Donald Trump (Republican candidate), at 51% and 48% of respondents answering respectively. This may be due to respondent bias, as the Nationscape dataset itself was in partnership with the Democracy Fund, it is expected that a greater number of respondents who support the democratic party would respond.

The important thing to note in this side-by-side plot is that the distribution of some of the variables differ greatly between the survey data and the ACS data. For example, in the survey data, the age group "74 and above" appear to be underrepresented and "18 to 29 year olds" are over-represented. This justifies the use of post-stratification to ensure that our regression estimates are reliable and therefore that our predictions are more reflective of the population distributions.

@@

Since we are modelling the results of the popular vote for the US election between two candidates, we ran a logistic regression using glmer() from the “lme4” package in R on the data from the Democracy Fund + UCLA Nationscape poll. We used a binary outcome of 1 as a vote for Donald Trump and the Republican Party and 0 as a vote for Joe Biden and the Democratic Party. This allows us to interpret the regression output as a given individual or stratum’s probability of voting for the Republican Party, and may prove more accurate than a linear model where outcomes may fall outside of the range 0 to 1 and be meaningless in the context of election results. The model specification was:

$$ P(Republican Win) = logit^{-1}(\delta_0  +  \delta_1gender_i + \delta_2age_i + \delta_3education_i + \delta_4employment_i +\delta_5(1|state_i))$$

As mentioned previously, there were a limited number of corresponding variables between the survey and post-stratification data sets. Of the variables selected, region was excluded from all models due to its direct relation to an individual’s state, a perhaps more significant and accurate factor in voter preference. Our final model includes state at a group level, age, gender, race, and employment status as factors, and education as a numeric linear variable. All coefficients in this model are significant at a level of 0.05, except for the coefficient on the employment level “unemployed”. Education was included as a numeric vector (despite its categorical levels) as it was insignificant within the model when inputted as a factor. This was also due to the cumulative nature of educational attainment and therefore its potential to be represented as a linear value (compared to something like race, for example). States entered the model as a group level effect to limit the number of individual coefficients in the model output and maintain overall model simplicity while still accounting for its influence. Age entered the model as a factor to better represent the differences between age groups, as voter preference may change non-linearly with age. It is also unlikely that voter preference shifts year to year by age, and thus does not need to be modeled as such.

There are a number of reasons for excluding household_income from this model beyond overall model diagnostics. Firstly, the UCLA survey data represents data about household income, while the variable FTOTINC representing family income was selected as the corresponding variable from the ACS data. While these may represent similar overall influences in the model, we cannot be sure that these variables would match on a respondent level and may thus compromise the integrity of the model if included. Furthermore, there are many “NA” values in both datasets for the household_income variable which may skew the model and post-stratification results. Accounting for income also increased the number of data cells by a factor of 24, greatly increasing the complexity of the data and model itself and decreasing the size of each individual cell. To account for economic factors in the model, employment was kept in the final model despite one of its coefficients being insignificant at p=0.05.

Two other models were attempted prior to our final model, both excluding household income and employment and differing by treatment of age (FIGURE #). Both of these models failed to converge and were thus discarded.

Post-stratification is used to properly balance the representation of variables across the target population to gain more precise estimates and create greater confidence in inferences being made. Post-stratification involves classifying each member of a population into a single subgroup or stratum to then calculate the probability of a particular outcome for each distinct stratum using a regression model. Because of this, post-stratification cannot be used in studies with observations that overlap or are not clearly classified, as this may inaccurately reflect the population. Another challenge is to find definitive lists of variables for an entire population that fall in line with the variables collected for the original sample. This may restrict the potential models and strata used for post-stratification analysis.

Based on the variables selected (51 levels of state, six levels of education, five levels of age, two levels of gender, seven levels of race, and three levels for employment), there were a potential 64,260 data cells with 37,530 populated by the ACS data.

@@

Figure 1 depicts the estimated outcome of Republican voters in each state. We can see rocky mountain states in the Midwest are all likely to vote Republican. Overall, after tallying up the probabilities of each state, the popular vote is in Demoratic favour of 51% to 49%. However, presidential elections in the American electoral college system do not follow the popular vote causing votes of different states to carry unequal weighting; this is where the impact of swing states really shine. Notable swing states are Pennsylvania, North Carolina, and Florida (Vox, 2020) where the votes lean Republican, Demoratic, and split respectively. This means that while we have estimates for the popular vote, we are unable to predict the outcome of the election with such close values.

In Figure 2, the grey dots represent the raw survey data while the black dots represents values post-stratification. We can see that after post-stratification, the three states to hold the largest votes for Republicans are West Virginia, Tennessee, and Idaho collecting about 60% proportion of the votes in favour of Trump. The strongest Democratic states are Massachusetts, Connecticut, and District of Columbia where only 40 percent of voters are Republican. As expected Florida, Ohio, and North Carolina gravitate to the 50-50 line as they are known to be swing states.

Figure 3 shows a strong positive correlation with age and the proportion of Republican voters both from the post-stratification data (the black solid line).  At ages 30-44 and 60-74, the proportion of voters are evenly divided and the youngest age group has extremely high proportions of Democratic votes reaching approximately 65%, while the opposite effect is observed for respondents ages 74 and over (though only reaching 57% Republican proportions). In contrast, the raw survey data (grey line) had the highest proportion of repoblican voters in ages 18-29 and 45 to 59 and substantially low republican proportions at ages 30-44 (32%).

Figure 4, shows a similar correlation between the raw survey data (grey dotted line), and post-stratified data (black line) between race/ethnicity to make up the Republican proportion of the popular vote. It is noted that Native American respondents are most likely to make up this proportion, with a strong correlation representing the highest proportion of republican voters (60%) , followed by White respondents with a republican proportion of approximately(55%). This is in contrast to the lowest proportion of republican voters seen in Black respondents, with a proportion of less than 10%.

Figure 5, depicts predicted Republican voting based on the respondent’s educational attainment. It is noted that in the raw survey data (grey dotted line), respondents who have achieved an education level less than high school, would have the weakest correlation to vote Republican. This is in contrast to the raw data seen among respondents who have attained a postgraduate degree, who have the highest correlation to vote Republican. These two ends of the categorical spectrum for educational attainment is seen to be the opposite in the post-stratified data (black solid line), with respondents having the lowest educational attainment (less than high school) having the highest correlation, and those with the highest educational attainment (post graduate degree), having the lowest correlation to vote Republican. We observe that in the post stratified data, there exists a decreasing trend in correlation to vote Republican as the level of education attainment increases.

Figure 6 depicts voters employment status. The graph shows that after post-stratification (black solid line), those who are employed have a 52% Republican predicted proportion. However, this value drops almost 10 points to a 42% Republican proportion to unemployed voters. On the other hand, the raw survey data (dotted grey line) shows the larger voter  proportion will be Democratic no matter a voter's employment status.

Figure 7 observes likelihood to vote Republican based on gender. Between the raw survey data, represented by the grey dotted line, and post stratified data, represented by the black solid line, there are similar trends seen. It is observed that females are less likely to vote Republican due to the lower correlation, as opposed to males who have a higher correlation to vote Republican.

@@

Given the bias in the survey, discussed in the Data section, we used MRP to adjust our prediction to better represent the eligible voting population of the US. While this would address some aspects of the bias, it cannot address all of it. For instance, we simply cannot account for the impact that COVID-19 as well as certain socio-economic conditions would have on the voter turnout. Voter turnout in America is historically low compared to other large democracies. With the postal ballot system in place for the 2020 elections, it is unclear how that would affect the demographic makeup of the voters who actually participate in the elections. The MRP, in the way that we adopted, cannot take into account voter participation. We simply assumed that there would be a 100% turnout, even though this is a highly unrealistic assumption. We expect voter turnout to be close to 50%, as in the past, and it is unclear how different segments of society would be represented in the final vote.

Our model assumed people would either vote for Trump or Biden. In reality, independent candidates would also win some of the votes. So a multinomial logistic regression model would have been more appropriate than a binary logistic regression model. In future work, we would take this into account.

@@

Bates, D., Maechler, M., Bolker, B., Walker, S. (2015). Fitting Linear Mixed-Effects Models Using lme4. Journal of Statistical Software, 67(1), 1-48. doi:10.18637/jss.v067.i01.

Bergsma, T. (2018). latexpdf: Convert Tables to PDF or PNG. R package version 0.1.6. https://CRAN.R-project.org/package=latexpdf

@@

Di Lorenzo, P. (2020). usmap: US Maps Including Alaska and Hawaii. R package version 0.5.1. https://CRAN.R-project.org/package=usmap

@@

Eddelbuettel, D. and Gilligan, J. (2020). tint: 'tint' is not 'Tufte'. R package version 0.1.3. https://CRAN.R-project.org/package=tint

@@

Kennedy, C., Blumenthal, M., Clement, S., Clinton, J. D., Durand, C., Franklin, C.,

McGeeney, K., Miringoff, L., Olson, K., Rivers, D., Saad, L. Witt, G. E., & Wlezien, C. (2018). An evaluation of the 2016 election polls in the United States. *Public Opinion Quarterly*, 82(1), 1-33.

@@

Larmarange, Joseph (2020). labelled: Manipulating Labelled Data. R package version 2.7.0. https://CRAN.R-project.org/package=labelled

@@

Vox (2020). The Electoral College, explained. https://youtu.be/ajavsMbCapY

Wickham, H. and Miller, E. (2019). haven: Import and Export ’SPSS’, ’Stata’ and ’SAS’ Files. R package version 2.1.1.

@@

@@

Zhu, H. (2020). kableExtra: Construct Complex Table with 'kable' and Pipe Syntax. R package version 1.2.1. https://CRAN.R-project.org/package=kableExtra
